Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-07-12 AnalyticStructure_main.tex
last before arxiv

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\Liouv}{\sop{\mathcal L}}
\newcommand{\Liouvnot}{\sop{\mathcal L_0}}
\newcommand{\coupl}{\sop{\mathcal K}}
\newcommand{\honetmp}[1][]{\op{H_1}\argument{#1}} 
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}
\newcommand{\fscd}[1]{\ket{f_p^\Phi(\omega)^{#1}}}

@@

\begin{abstract}
By performing a critical analysis of the fundamental equations of linear-response  (LR) formalisms in molecules, 
we explore the interplay between locality of the response density operator and  numerical convergence of LR-related quantities.
%We show that the first ionization potential of the system behaves as a threshold energy: for frequencies below this value, it is possible to parametrize the response density by employing localized states only.
We show that for frequencies below the first ionization potential (IP) of the system, it is possible to parametrize the response density by employing localized states only.
Above this threshold energy, such a locality property cannot be achieved. 
Such considerations may be transposed in terms of the molecule's excited states. We show that not all the system's excitations can be considered on equal footing.
There is a discrete sector of excitations -- which may extend above IP -- that can be parametrized by observable, localized states, which can be computationally expressed with high precision, provided an adequate level of completeness. 
We present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  
The remaining excitation modes belong to a continuum spectrum that, on the contrary, is not directly associated to observable properties and can only be effectively represented in a given computational setup.
Such considerations are important not only for reproducibility of the results among different computer codes 
employing diverse formalisms, but also in view of providing a deeper understanding on the impact of models' 
approximations on the scientific outcomes of the simulation.
\end{abstract}

@@

\section{Introduction}
Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the 
computational reliability of the results. More specifically, a computational result may be considered reliable for various reasons.
It may be, for example, in agreement with some previous observation or measurement that the employed computational model is susceptible to reproduce.
The model itself may in this way be classified as ``accurate''; it can then be employed to extract, \textit{in silico},
other quantities that might be then used to \emph{predict} or \emph{interpret} experiments.
Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
that a given computational model provides for a specific set of quantities, regardless of their potential agreement with 
experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertainties
of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
employing the same model can be compared.

@@

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed 
for ground state calculations of molecules. 
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behavior of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will also reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in the Appendix.

@@

Let us start our discussion by revisiting the impact that the Isolated BC have on the 
solutions of the time-independent Schr\"odinger Equation.
We here present well-known concepts of elementary quantum mechanics, under a perspective that
will turn out to be very useful in the forthcoming considerations about linear-response quantities.

Consider a (one-body) wavefunction $\ket{\psi}$ solution of the equation
$\op H\ket{\psi}=\eps\ket{\psi}$, where we split the Hamiltonian operator into the sum of
a kinetic $\op T$ and a potential $\op V \equiv \op H - \op T$ term.
In view of analysing the impact of the BC, it is instructive to transform 
the Schr\"odinger equation in a scattering problem, by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \op G_H(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\eps) = (\eps-\op T)^{-1}$,
resolvent of the free-particle Hamiltonian $H_s \equiv \op T$.

@@

This fact is made apparent by the expression of 
the kernel of $\op G_H(\eps)$ in the coordinate representation. 
When dealing with Isolated BC, it is easy to see that the solution of Eq.~\eqref{SEHelmholtz1}
can be expressed by:
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\eps) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{-\alpha|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; \eps  < 0\; \\ 
\frac{e^{\ii \alpha |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; \eps \geq 0
\end{cases} \;,
\ee
where $\alpha = \sqrt{2|\eps|}$.

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;.
\ee
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part $\Liouvnot$ plus a coupling term $ \coupl$. The action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$ encodes the response of the $\dm$-dependent potential to a modification of the density operator.
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidences that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$.
An operator $\op O_\perp$ with this feature satisfy the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\op O_\perp \equiv 
\dmnot\op O \op Q_0 + \op Q_0 \op O\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
In other terms, the condition $\dm'_\Phi = \dm'_{\Phi\,\perp}$ holds.

We can therefore introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states (FS) are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p} 
$ as a source term.

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} = 
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
where the bound state character of $\{\bra{\psi_p}\op O\}$ behaves as a regulator.

@@

\subsection{The analytic structure of linear susceptibility}
Each excitation ``mode'' of the system $\op E_a$ with energy $\Omega_a$, can be associated to a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$,
entirely defined in the unoccupied subspace. 
In this way we can parametrize, without loss of generality, the excitation operators of the system as:
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
For each choice of $\{\phi^a_p,\chi^a_p\}_{p \in a}$, these objects represent, respectively, the state in which $\ket{\psi_p}$
is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$.
We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2}
is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry 
of the excitation.
Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues 
$\Omega_a \rightarrow -\Omega_a$ and the associated excitation with negative energy are described by the transpose pair $\{\chi^a_p,\phi^a_p\}$.
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

With the same spirit of the previous sections, we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
\ket{\phi^a_p} &= \op G_H(\Omega_a+\eps_p)\left(\op V \ket{\phi^a_p} + \ket{s_p[\op E_a]} \right) \;, \nn \\
\bra {\chi^a_p} &=\left(\bra{\chi^a_p} \op V  + \bra{s_p[\op E_a]}\right)  \op G_H(-\Omega_a+\eps_p)\;,
\end{align}
where the source terms are here defined as:
\be
 \ket{s_p[\op O]} =  \op Q_0 \op V'[\op O] \ket{\psi_p}\,, \quad
 \bra{s_p[\op O]} =   \bra{\psi_p} \op V'[\op O] \op Q_0 \nn \;.
\ee
We first point out that the Helmholtz kernel associated to the $\ket{\chi^a_p}$ state 
contains an exponential damping factor for \emph{all} the positive values of $\Omega_a$.
The asymptotic long-range behavior of $\ket{\chi^a_p}$ thus depends only on the locality properties of 
$\bra{s_p[\op E_a]}$, which in turn is related to the behavior of the operator $\op V'[\op E_a]$.

On the other hand, the long-range behavior of $\op G_H(\Omega_a+\eps_p)$ depends from 
$\Omega_a$, with a threshold of value $\Omega_a=|\eps_p|$. 
For energies above this value, the behavior of the state $\ket{\phi^a_p}$ is 
delocalized regardless of the particular nature of the operators $\op V$ and $\op V'$.
This means that, if an excitation operator $\op E_a$ has an energy such that
$\Omega_a>|\eps_p|$  for at least one of the $p \in a$ defining the excitation,
such operator \emph{cannot} be parametrized by employing localized states only.

@@

The susceptibility functional has therefore a peculiar analytic structure, that
is made of multiple threshold energies, one for each of the ionization potentials
associated to the occupied states. 
If an excitation $\op E_a$ has a energy $\Omega_a < |\eps_a|$ with $\eps_a = \mathrm{max}\left(\eps_p\right)_{p\in a}$,
it may belong to a discrete spectrum; otherwise it \emph{has} to belong to a continuum of states. 
The typical structure of the excitations landscape is depicted in figure \ref{ExcitationLandscape}, once again assuming a reasonable 
behavior of the perturbed potential $\op V'$.
Below the first ionization energy $|\eps_h|$ only localized excitations are present. Conversely, in the intermediate energy interval that extends from $|\eps_h|$ to the 
absolute value of the deepest occupied orbital, other discrete excitations may exist, and are embedded in a continuum of delocalized excitations. 
For energies higher than $|\eps_1|$ \emph{no localized} excitations are anymore possible,
regardless of the particular behavior of $\op V'$: the excitation landscape in this region only contains continuum states.

@@

The analysis of the above section reveals that the excitation spectrum is composed by two distinct sectors. 
The first one is realized by the (finite) set of localized excitations with discrete eigenvalues. 
Borrowing the terminology introduced in section \ref{SEopenSystem} to the present case, we can affirm that such localized excitations behaves as
\emph{observable} quantities and the associated energies possess ideal features of reproducibility:
if the numerical basis employed is complete enough to
represent the states $\{\phi_p^a,\chi_p^a\}$, the energy $\Omega_a$ 
converges to its reference value.

The second sector contains a continuum spectrum of delocalized excitations. The associated states $\ket{\phi_p^a}$ behave as the unbound virtual orbitals analyzed in \cite{boffi2016} 
and are deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and 
the associated spectrum exhibit an explicit dependence from the size of the domain. 
Excitations belonging to this sector thus lose their intrinsic physical meaning and have to be understood simply as
a representation of the density of states (DOS) of the Liouvillian superoperator.

@@

By analysing the fundamental equations of LR calculations of molecules, we have shown that the response density for values of $\omega$ above the first IP
\emph{have} to be expressed in terms of genuinely delocalized states. 
Below this threshold,
under reasonable assumptions on the perturbing operators,
the perturbed density operator is genuinely localized.
In this localized regime, it is possible to achieve convergence of the numerical result
by increasing the completeness of the basis with the usual GS rationales.
One might expect that more complete basis will give more precise results, regardless of their asymptotic behavior.
A comparison among various codes in view of reproducibility should therefore be performed 
initially in this regime.

>>>>

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\Liouv}{\sop{\mathcal L}}
\newcommand{\Liouvnot}{\sop{\mathcal L_0}}
\newcommand{\coupl}{\sop{\mathcal K}}
\newcommand{\honetmp}[1]{\op{H_1}\argument{#1}} 
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}
\newcommand{\fscd}[1]{\ket{f_p^\Phi(\omega)^{#1}}}
\newcommand{\GH}{\op G_{\op T}}

@@

\begin{abstract}
By performing a critical analysis of the fundamental equations of linear-response (LR) formalism in molecules, 
we explore the interplay between locality of the response density operator and  numerical convergence of LR-related quantities.
We show that for frequencies below the first ionization potential (IP) of the system, it is possible to express the response density by employing localized states only.
Above this threshold energy, such a locality property cannot be achieved. 
Such considerations may be transposed in terms of the molecule's excited states. We show that not all the system's excitations can be considered on equal footing.
There is a discrete sector of excitations -- which may extend above IP -- that can be parametrized by observable, localized states, which can be computationally expressed with high precision, provided an adequate level of completeness. 
We present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  
The remaining excitation modes belong to a continuum spectrum that, on the contrary, is not directly associated to observable properties and can only be effectively represented in a given computational setup.
Such considerations are important not only for reproducibility of the results among different computer codes 
employing diverse formalisms, but also in view of providing a deeper understanding on the impact of models' 
approximations on the scientific outcomes of the simulation.
\end{abstract}

@@

\section{Introduction}
Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the 
computational reliability of the results. More specifically, a computational result may be considered reliable for various reasons.
For example, it may be in agreement with some previous observation or measurement that the employed computational model is susceptible to reproduce.
The model itself may in this way be classified as ``accurate''; it can then be employed to extract, \textit{in silico},
other quantities that might be then used to \emph{predict} or \emph{interpret} experiments.
Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
that a given computational model provides for a specific set of quantities, regardless of their potential agreement with 
experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertainties
of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
employing the same model can be compared.

@@

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed 
for ground state calculations of molecules. 
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behavior of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will also reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in Appendix~\ref{compdetails}.

@@

Let us start our discussion by revisiting the impact that the Isolated BC have on the 
solutions of the time-independent Schr\"odinger Equation.
We here present well-known concepts of quantum mechanics, under a perspective that
will turn out to be very useful in the forthcoming considerations about linear-response quantities.

Consider a (one-body) wavefunction $\ket{\psi}$ solution of the equation
$\op H\ket{\psi}=\eps\ket{\psi}$, where we split the Hamiltonian operator into the sum of
a kinetic $\op T$ and a potential $\op V \equiv \op H - \op T$ term.
In view of analysing the impact of the BC, it is instructive to transform 
the Schr\"odinger equation in a scattering problem, by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \GH(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\GH(\eps) = (\eps-\op T)^{-1}$,
resolvent of the free-particle Hamiltonian $H_s \equiv \op T$.

@@

This fact is made apparent by the expression of 
the kernel of $\GH(\eps)$ in the coordinate representation. 
When dealing with Isolated BC, it is easy to see that the solution of Eq.~\eqref{SEHelmholtz1}
can be expressed by:
\be\lb{HelmholtzKernelDef1}
\bra{\r} \GH(\eps) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{-\alpha|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; \eps  < 0\; \\ 
\frac{e^{\ii \alpha |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; \eps \geq 0
\end{cases} \;,
\ee
where $\alpha = \sqrt{2|\eps|}$.

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;.
\ee
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part $\Liouvnot$ plus a coupling term $ \coupl$. The action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$ encodes the response of the $\dm$-dependent potential to a modification of the density operator.
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidences that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$.
An operator $\op O_\perp$ with this feature satisfy the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\op O_\perp \equiv 
\dmnot\op O \op Q_0 + \op Q_0 \op O\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in empty subspace of $\hnot$
(for more details see Appendix~\ref{LiouvillianAction}).
In other terms, the condition $\dm'_\Phi = \dm'_{\Phi\,\perp}$ holds.

We can therefore introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states (FS) are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \GH(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op \Phi(\omega) +\op V'[\dm'_\Phi](\omega) )\ket{\psi_p} 
$ as a source term.

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} = 
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule 
perturbed by a static electric field. 
Care has been taken in keeping the field strenghts $F_j$ small enough to 
preseve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
where the bound state character of $\{\bra{\psi_p}\op O\}$ behaves as a regulator.
This is naturally achieved in the above mentioned Krylov space treatments, which
explicitly deal with the expression of the scalar products. However
it is important that the underlying basis set is able to express delocalised states.

@@

\subsection{The analytic structure of linear susceptibility}
Each excitation ``mode'' of the system $\op E_a$ with energy $\Omega_a$, can be associated to a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$,
entirely defined in the unoccupied subspace. 
In this way we can parametrize, without loss of generality, the excitation operators of the system as:
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
For each choice of $\{\phi^a_p,\chi^a_p\}_{p \in a}$, these objects represent, respectively, the state in which $\ket{\psi_p}$
is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$.
We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2}
is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry 
of the excitation.
Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues 
$\Omega_a \rightarrow -\Omega_a$ and the associated excitation with negative energy are described by the transposed pair $\{\chi^a_p,\phi^a_p\}$.
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

With the same spirit of the previous sections, we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
\ket{\phi^a_p} &= \GH(\Omega_a+\eps_p)\left(\op V \ket{\phi^a_p} + \ket{s_p[\op E_a]} \right) \;, \nn \\
\bra {\chi^a_p} &=\left(\bra{\chi^a_p} \op V  + \bra{s_p[\op E_a]}\right)  \GH(-\Omega_a+\eps_p)\;,
\end{align}
where the source terms are here defined as:
\be
 \ket{s_p[\op O]} =  \op Q_0 \op V'[\op O] \ket{\psi_p}\,, \quad
 \bra{s_p[\op O]} =   \bra{\psi_p} \op V'[\op O] \op Q_0 \nn \;.
\ee
We first point out that the Helmholtz kernel associated to the $\ket{\chi^a_p}$ state 
contains an exponential damping factor for \emph{all} the positive values of $\Omega_a$.
The asymptotic long-range behavior of $\ket{\chi^a_p}$ thus depends only on the locality properties of 
$\bra{s_p[\op E_a]}$, which in turn is related to the behavior of the operator $\op V'[\op E_a]$.

On the other hand, the long-range behavior of $\GH(\Omega_a+\eps_p)$ depends from 
$\Omega_a$, with a threshold of value $\Omega_a=|\eps_p|$. 
For energies above this value, the behavior of the state $\ket{\phi^a_p}$ is 
delocalized regardless of the particular nature of the operators $\op V$ and $\op V'$.
This means that, if an excitation operator $\op E_a$ has an energy such that
$\Omega_a>|\eps_p|$  for at least one of the $p \in a$ defining the excitation,
such operator \emph{cannot} be parametrized by employing localized states only.

@@

The susceptibility functional has therefore a peculiar analytic structure, that
is made of multiple threshold energies, one for each of the ionization potentials
associated to the occupied states. 
If an excitation $\op E_a$ has an energy $\Omega_a < |\eps_a|$ with $\eps_a = \mathrm{max}\left(\eps_p\right)_{p\in a}$,
it may belong to a discrete spectrum; otherwise it \emph{has} to belong to a continuum of states. 
The typical structure of the excitation landscape is depicted in figure \ref{ExcitationLandscape}, once again assuming a reasonable 
behavior of the perturbed potential $\op V'$.
Below the first ionization energy $|\eps_h|$ only localized excitations are present. Conversely, in the intermediate energy interval that extends from $|\eps_h|$ to the 
absolute value of the deepest occupied orbital, other discrete excitations may exist, and are embedded in a continuum of delocalized excitations. 
For energies higher than $|\eps_1|$ \emph{no localized} excitations are anymore possible,
regardless of the particular behavior of $\op V'$: the excitation landscape in this region only contains continuum states.

@@

The analysis of the above section reveals that the excitation spectrum is composed by two distinct sectors. 
The first one is realized by the (finite) set of localized excitations with discrete eigenvalues. 
Borrowing the terminology introduced in section \ref{SEopenSystem} to the present case, we can affirm that such localized excitations behave as
\emph{observable} quantities and the associated energies possess ideal features of reproducibility:
if the numerical basis employed is complete enough to
represent the states $\{\phi_p^a,\chi_p^a\}$, the energy $\Omega_a$ 
converges to its reference value.

The second sector contains a continuum spectrum of delocalized excitations. The associated states $\ket{\phi_p^a}$ behave as the unbound virtual orbitals analyzed in \cite{boffi2016} 
and are deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and 
the associated spectrum exhibits an explicit dependence from the size of the domain. 
Excitations belonging to this sector thus lose their intrinsic physical meaning and have to be understood simply as
a representation of the density of states (DOS) of the Liouvillian superoperator.

@@

By analysing the fundamental equations of LR calculations of molecules, we have shown that the response density for values of $\omega$ above the first IP
\emph{have} to be expressed in terms of genuinely delocalized states. 
Below this threshold,
under reasonable assumptions on the perturbing operators,
the perturbed density operator is genuinely localized.
In this localized regime, it is possible to achieve convergence of the numerical result
by increasing the basis completeness in the same spirit of GS calculations.
One might expect that more complete basis will give more precise results, regardless of their asymptotic behavior.
A comparison among various codes in view of reproducibility should therefore be performed 
initially in this regime.
